---
title: "QAC380-data"
author: "Jiner Zheng"
date: "3/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r load packages and data}
library(dplyr)
RiskDaTa <- read.csv("Data /RiskData.csv")
OffData <- read.csv("Data /OffenseData.csv")
DrugData <- read.csv("Data /DrugData.csv")
AdressDaTa <- read.csv("Data /ClientAddressData.csv") %>% distinct() # remove redundant rows from adress data
ClientDaTa <- read.csv("Data /ClientData.csv")
EpisodeDaTa <- read.csv("Data /EpisodeData.csv")
IncidentDaTa <- read.csv("Data /IncidentData.csv")
TraumaDaTa <- read.csv("Data /EpisodeTraumaData.csv") # need to check
LsiData <- read.csv("Data /LsiData.csv")
```

```{r}
#FOR LEAH
setwd("~/Desktop/QAC 380- Stat. Consulting")

RiskDaTa <- read.csv("RiskData.csv")
OffData <- read.csv("OffenseData.csv")
DrugData <- read.csv("DrugData.csv")
AdressDaTa <- read.csv("ClientAddressData.csv") %>% distinct() # remove redundant rows from adress data
ClientDaTa <- read.csv("ClientData.csv")
EpisodeDaTa <- read.csv("EpisodeData.csv")
IncidentDaTa <- read.csv("IncidentData.csv")
TraumaDaTa <- read.csv("EpisodeTraumaData.csv") # need to check
LsiData <- read.csv("LsiData.csv")


```

### Risk Data: Main VARS: SuicideRisk_Total / HomicideRiskTotal
```{r Risk Data}
# Aggregated new variable: Total Suicide Risk Level & Total Homicide Risk Level
RiskDaTa <- RiskDaTa %>% 
  mutate(SuicideRisk_Total = ThoughtBetterOffDeadInTheLastMonth + WantedToHarmSelfInTheLastMonth + SuicidalThoughtsInTheLastMonth + SuicidePlanInTheLastMonth + AttemptedSuicideInTheLastMonth + FeltHopelessAboutFutureInTheLastMonth + AttemptedSuicideInLifeTime + HasFamilyMembersAttemptedOrCompletedSuicideInLifetime,
         HomicideRisk_Total = ThoughtAboutHurtingOthersInTheLastMonth +
  ThouthAboutHutingStaffInTheLastMonth + PannedToKillSomeoneElseInTheLastMonth + 
  TriedToKillSomeoneEliseInTheLastMonth + EverKilledSomeoneInLifetime +
  EverHurtAuthorityFigureInLifetime + EverHurtAuthorityFigureInLifetimeUsedWeapon)
```
#### latested_risk_by_client: data frame of most recently assessed risk levels
variables: StudyClientId, AssessmentYear, AssessmentType, SuicideRisk_Total, HomicideRiskTotal
```{r risk level by most recent year}
# for merge purpose
latest_risk_by_client <- RiskDaTa %>% 
  group_by(StudyClientId) %>% 
  arrange(AssessmentYear) %>% 
  summarise(SuicideRisk_Total = last(SuicideRisk_Total),
            HomicideRisk_Total = last(HomicideRisk_Total),
            AssessmentYear = last(AssessmentYear),
            AssessmentType = last(AssessmentType))
```


```{r assessment methods for risk data}
# assessment methods year
# RiskDaTa %>% group_by(AssessmentType) %>% 
#  summarise(min_year = min(AssessmentYear),
#            max_year = max(AssessmentYear),
#            number_of_clients = n_distinct(StudyClientId)) # unique client counts
```

### Offense Data: Main VAR (NO NEED TO RUN THIS SECTION)

Class A: most severe (e.g.: murder, threatening, sexual assault)
```{r Offense Data}
OffData <- OffData %>% 
  mutate(ChargeSeverity = case_when(SentencedChargeOffenseClass == "A" ~ 5,
                                    SentencedChargeOffenseClass == "B" ~ 4,
                                    SentencedChargeOffenseClass == "C" ~ 3,
                                    SentencedChargeOffenseClass == "D" ~ 2,
                                    SentencedChargeOffenseClass == "U" ~ 1))
```

Take a look at missing values of ChargeSeverity --> Non-labeled offense calss
```{r}
OffData %>% filter(is.na(ChargeSeverity)) %>%
  group_by(SentencedChargeDescription) %>% 
  summarise(count=n())
```
#### TO DO: do we need to assign a new class to these offenses or do we just treat it as a binary variable instead (have offense history = 1, not have = 0)?
#### NOTE: the Offense Data DOES NOT have ClientID or any thing that could be used to reference another  dataset. We might need to discard it.

### Alcohol Drug Problem & Criminal History Data
```{r Drug use data}
# use AlcoholDrugProblem from LsiData instead
latest_drug_by_client <- LsiData %>% 
  group_by(StudyClientId) %>% 
  arrange(AssessmentYear) %>% 
  summarise(AlcoholDrugProblems = last(AlcoholDrugProblems),
            AssessmentYear = last(AssessmentYear),
            AssessmentType = last(AssessmentType))

latest_criminal_by_client <- LsiData %>% 
  group_by(StudyClientId) %>% 
  arrange(AssessmentYear) %>% 
  summarise(CriminalHistory = last(CriminalHistory),
            AssessmentYear = last(AssessmentYear),
            AssessmentType = last(AssessmentType))
```

### Incident Data: Main VAR: IncidentScore (from Incident_byclient dataset)
#### IncidentScore: a composite measure based on the number of incidents corresponding to its level of severity (i.e. the sum of count * incident level)
#### IncidentLevel: discrete values 1,2,5 according to its type(5=most severe, i.e. [I] level incident)
All types of incidents are:
```{r incident types}
# frequency table of incident type
descr::freq(IncidentDaTa$IncidentType,plot=FALSE)
```

```{r Incident data}
# split client Ids and recode incident type by severity
IncidentDaTa2 <- IncidentDaTa %>% 
  mutate(IncidentLevel = case_when(grepl("\\[I]", IncidentType) ~ 5, 
                                   grepl("\\[II]", IncidentType) ~ 2, 
                                   grepl("\\[III]", IncidentType) ~ 1),
    StudyClientID = strsplit(as.character(StudyClientIDs),"\\|")) %>% 
  # grepl() --> detect if strings contain certain pattern
  # strsplit() --> split strings by certain character
  # double \\ to interpret | symbol literally
  # NOTE: can't split study episodes ID becuz some clients have more than 1 episodes
  tidyr::unnest(StudyClientID) %>% 
  mutate(StudyClientID=as.numeric(StudyClientID)) # change ID back to numeric

Incident_by_client <- IncidentDaTa2 %>% 
  group_by(StudyClientID, IncidentType, IncidentLevel) %>% 
  summarize(count_by_incident = n(), .groups="keep") %>% # get number of incidents by type and client
  group_by(StudyClientID) %>% 
  summarize(IncidentScore = sum(IncidentLevel * count_by_incident),
  # the following three vars are just for checking the calculation was correct
            IncidentLevels = paste(IncidentLevel, collapse = " & "),
            IncidentCounts = paste(count_by_incident, collapse = "&"),
            IncidentTypes = paste(IncidentType, collapse = " & "))
names(Incident_by_client)[1] <- "StudyClientId"
```


### Trauma Data
```{r Trauma DaTa}
# recode trauma variables True -> 1 & False -> 0
TraumaDaTa2 <- TraumaDaTa %>% 
  mutate(DomesticViolenceHistory = if_else(DomesticViolenceHistory == "True", 1,0),
         PhysicalAbuseVictim = if_else(PhysicalAbuseNarrative == "True", 1,0),
         SexualAbuseVictim = if_else(SexualAbuseVictim == "True",1,0),
         OtherTraumaHistory = if_else(OtherTraumaHistory == "True",1,0),
         EverDiagnosedWithPTSD = if_else(EverDiagnosedWithPTSD == "True",1,0),
         HasChronicPainProblems = if_else(HasChronicPainProblems == "True",1,0),
         TraumaHistory = case_when(DomesticViolenceHistory == 1 | PhysicalAbuseVictim == 1 | SexualAbuseVictim == 1 | OtherTraumaHistory == 1 | EverDiagnosedWithPTSD == 1 | HasChronicPainProblems == 1 ~ 1, DomesticViolenceHistory == 0 | PhysicalAbuseVictim == 0 | SexualAbuseVictim == 0 | OtherTraumaHistory == 0 | EverDiagnosedWithPTSD == 0 | HasChronicPainProblems == 1 ~ 0))

# whether or not client has any kind of trauma 
Trauma_by_client <- TraumaDaTa2 %>% 
  select(StudyClientId, TraumaHistory) %>% 
  distinct() %>% 
  group_by(StudyClientId) %>% 
  summarise(TraumaHistory=max(TraumaHistory))
```

### ClientAllData: merged ClientData with Risk / Incident / Alcohol&Drug / Criminal History & Trauma & Address datasets
```{r merging data}
# merge all data frames together
ClientAllData <- ClientDaTa %>% 
  left_join(latest_criminal_by_client[,1:2], by="StudyClientId") %>%
  left_join(latest_drug_by_client[,1:2], by="StudyClientId") %>% 
  left_join(latest_risk_by_client[,1:3], by="StudyClientId") %>% 
  left_join(Incident_by_client[,1:2], by="StudyClientId") %>% 
  left_join(Trauma_by_client, by="StudyClientId") %>% 
  left_join(AdressDaTa %>% group_by(StudyClientId) %>% 
              summarise(AddressType = last(AddressType),
                        City = last(City),
                        State = last(State),
                        Zip = last(Zip)), by="StudyClientId")

# recode NA in the IncidentScore to be 0
ClientAllData$IncidentScore[is.na(ClientAllData$IncidentScore)] <- 0

ClientAllData %>%
  filter(CurrentAge >= 18) -> ClientAllData

```

```{r}
library(ggplot2)

ggplot(data= na.omit(subset(ClientAllData, select = c(CriminalHistory))),
       aes(x=reorder(as.character(CriminalHistory), CriminalHistory), fill = as.character(CriminalHistory)))+
  geom_bar(na.rm = TRUE)+
  ggtitle("Clients with Criminal History")+
  xlab("Criminal History")

```
```{r}
ggplot(data= na.omit(subset(ClientAllData, select = c(AlcoholDrugProblems))),
       aes(x=reorder(as.character(AlcoholDrugProblems), AlcoholDrugProblems), fill = as.character(AlcoholDrugProblems)))+
  geom_bar(na.rm = TRUE)+
  ggtitle("Clients with Alcohol and Drug Problems")+
  xlab("Alcohol and Drug Problems")
```

```{r}
ggplot(data= na.omit(subset(ClientAllData, select = c(SuicideRisk_Total))),
       aes(x=reorder(as.character(SuicideRisk_Total), SuicideRisk_Total), fill = as.character(SuicideRisk_Total)))+
  geom_bar(na.rm = TRUE)+
  ggtitle("Risk of Suicide in Clients")+
  xlab("Suicide Risk")
```
```{r}
ggplot(data= na.omit(subset(ClientAllData, select = c(HomicideRisk_Total))),
       aes(x=reorder(as.character(HomicideRisk_Total),HomicideRisk_Total), fill = as.character(HomicideRisk_Total)))+
  geom_bar(na.rm = TRUE)+
  ggtitle("Homicidal Risk of Clients")+
  xlab("Homicide Risk")
```

```{r}
ggplot(data=ClientAllData)+
 geom_density(aes(x=IncidentScore))+
 ggtitle("Client Incidents")
```

```{r}
ggplot(data= na.omit(subset(ClientAllData, select = c(TraumaHistory))),
       aes(x=reorder(as.character(TraumaHistory),TraumaHistory), fill = as.character(TraumaHistory)))+
  geom_bar(na.rm = TRUE)+
  ggtitle("Clients Assessed for Trauma")+
  xlab("Trauma")
```
```{r}
require(tidyverse)
top4cities <- ClientAllData %>%
    group_by(City) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

top4cities <- na.omit(top4cities)

ggplot(data = top4cities, aes(x = City, y = count, fill= City)) +
    geom_col()+
  ggtitle("Clients ")

```
```{r}
#Multiple Regression 
my.lm1 <- lm(IncidentScore~CriminalHistory+AlcoholDrugProblems+SuicideRisk_Total+HomicideRisk_Total+TraumaHistory, data= ClientAllData)
summary(my.lm1)

my.lm2 <- lm(IncidentScore~CriminalHistory+AlcoholDrugProblems+SuicideRisk_Total+HomicideRisk_Total+TraumaHistory+CurrentAge+as.factor(MaritalStatus)+as.factor(Race), data= ClientAllData)
summary(my.lm2)

```

```{r}
my.lm3 <- lm(IncidentScore ~ SuicideRisk_Total + CriminalHistory +
SuicideRisk_Total*AlcoholDrugProblems, data = ClientAllData)
summary(my.lm3)
```

