---
title: "Latent class analysis"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r load in Client Data}
library(dplyr)
library(plyr)
library(poLCA)
library(plotly)
library(ggplot2)
mydata <- read.csv('ClientAllData.csv')[-1]
mydata$City[mydata$City %in% c("Htfd","Hartford")] <- "Hartford"
mydata$City[mydata$City %in% c("Norwich","NORWICH")] <- "Norwich"
mydata$City[mydata$City %in% c("E Hartford","E Htfd","East Hartford")] <- "East Hartford"
mydata$City[mydata$City %in% c("middletown","Middletown")] <- "Middletown"
mydata$City[mydata$City %in% c("N Haven","New haven","New Haven","NEW HAVEN")] <- "New Haven"
mydata$City[mydata$City %in% c("New Brit","New Britain")] <- "New Britain"
mydata$City[mydata$City %in% c("W Haven","West Haven")] <- "West Haven"
mydata$City[mydata$City %in% c("waterbury","Waterbury","Wtby")] <- "Waterbury"
```

# Latent Class Analysis

#### 1. First, make sure all variables are factors and subset the data to include only LCA classification variables and covariates:

```{r subset data}
lca_subset<-subset(mydata, select=c(StudyClientId, age_group, Gender, Race, MaritalStatus, Religion, Aod1_class, Aod2_class, Disruption_class, LegalNonConform_class, LifeFunctionality_class, Suicide_class, Homicide_class, Trauma_class))
### recode missing entries as NA
lca_subset$Gender[!(lca_subset$Gender %in% c("M","F"))]<- NA
lca_subset$Race[lca_subset$Race %in% c("Not on file","Undisclosed")] <- NA
lca_subset$MaritalStatus[lca_subset$MaritalStatus %in% c("","Not Specified")] <- NA
lca_subset$Religion[lca_subset$Religion %in% c('','Unknown')] <- NA
### drop missing values
lca_subset <- lca_subset %>% tidyr::drop_na()
```

```{r recode VARS values to numbers}
# recode response values to numbers starting at "1" (It's a poLCA thing)
lca_subset$age_group<-revalue(lca_subset$age_group, 
                              c("18-30"="1", 
                                "31-50"="2",
                                ">50"="3"))
lca_subset$Gender<-revalue(lca_subset$Gender, c("M"="1", "F"="2"))
lca_subset$Race<-revalue(lca_subset$Race, 
                         c("Caucasian or White"="1", 
                           "African American or Black"="2",
                           'American Indian or Alaskan Native'='3',
                           'Asian' = '4', 
                           'Native Hawaiian/Other Pacific Islander'='5',
                           'Multi-Racial'='6',
                           'Some other race' = '7'))
lca_subset$MaritalStatus<-revalue(lca_subset$MaritalStatus, 
                                  c("Single/Never Married"="1", 
                                    "Married"="2",
                                    "Divorced/Annulled"='3',
                                    'Civil Union'='4',
                                    'Legally separated'='5',
                                    'Widow/widower'='6',
                                    'Other'='7'))
lca_subset$Religion<-revalue(lca_subset$Religion, 
                             c("Catholic"="1", 
                               "Atheist"="2",
                               "Agnostic"="3",
                               "Protestant"="4",
                               "Christian"="5",
                               "Islamic"="6",
                               "Baptist"="7",
                               "Pentecostal"="8",
                               "Buddhist"="9",
                               "Jewish"="10",
                               "Orthodox Christian"="11",
                               "Mormon"="12",
                               "None"="13",
                               "Other"="14"))
lca_subset$Aod1_class<-revalue(lca_subset$Aod1_class, 
                               c("low"="1", "medium"="2",'high'='3'))
lca_subset$Aod2_class<-revalue(lca_subset$Aod2_class, 
                               c("low"="1", "medium"="2",'high'='3'))
lca_subset$Disruption_class<-revalue(lca_subset$Disruption_class, 
                                     c("low"="1", "medium"="2",'high'='3'))
lca_subset$LegalNonConform_class<-revalue(lca_subset$LegalNonConform_class, c("low"="1", "medium"="2",'high'='3'))
lca_subset$LifeFunctionality_class<-revalue(lca_subset$LifeFunctionality_class, c(c("low"="1", "medium"="2",'high'='3')))
lca_subset$Suicide_class<-revalue(lca_subset$Suicide_class,
                                  c("low"="1", "medium"="2",'high'='3'))
lca_subset$Homicide_class<-revalue(lca_subset$Homicide_class, 
                                   c("low"="1", "medium"="2",'high'='3'))
lca_subset$Trauma_class<-revalue(lca_subset$Trauma_class, c("no/unknown"="2", "yes"="1"))
```

```{r factorize all VARS}
# make sure all variables are factors
lca_subset[-1] <- lapply(lca_subset[-1], factor)
```

#### 2. Second, define the LCA formula.

-   Variables in parentheses are the latent class classification variables.

-   Variables outside of the parentheses are covariates (not included in the LCA).

-   Finally, run the LCA specifying a range of classes

-   LCA: Classification VARS: ASUS / Risk/ Trauma

-   covariates: demographics (no location)

-   chisq+post-hoc: latent class vs. Outcome: Incident Score --\> binary

```{r define covariates and variables}
f <- cbind(Aod1_class,Aod2_class,Disruption_class,LegalNonConform_class,LifeFunctionality_class,Suicide_class,Homicide_class,Trauma_class)~age_group+Race+MaritalStatus+Religion
```

```{r check missing values}
sapply(lca_subset, function(x) sum(is.na(x)))
```

```{r run poLCA analysis}
# latent class analysis specifying 1-3 classes
lCA1 <- poLCA(f,lca_subset, nclass=1,nrep=15) 
lCA2 <- poLCA(f,lca_subset, nclass=2,nrep=15, graphs = T)
lCA3 <- poLCA(f,lca_subset, nclass=3,nrep=15, graphs = T)

```

```{r calculate entropy}
# Calculate entropy (3-class mode)l- values closer to 1.0 indicate greater separation of the classes.
entropy<-function (p) sum(-p*log(p))
error_prior <- entropy(lCA3$P) # Class proportions
error_post <- mean(apply(lCA3$posterior, 1, entropy))
LCA3_entropy <- (error_prior - error_post) / error_prior
LCA3_entropy
```

```{r predict class membership}
#predicted class membership is in:
lCA3$predclass[1:30]
#add variable to data set with all variables so it can be used as predictor variable:
lca_subset$class <- lCA3$predclass
```

```{r merge class back to orginal data, eval=FALSE}
# change eval to TRUE to run this chunk of code in html
# merge class membership variable with original data
data <- mydata %>% 
  inner_join(lca_subset[,c(1,15)], by = "StudyClientId")
write.csv(data,"client_class.csv")
```
