---
title: "LCA-followup"
author: "Jiner Zheng"
date: '2022-05-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load data and dependencies, include=FALSE}
library(dplyr) # data manipulation
library(knitr) # kable() --> beautify table
library(kableExtra) # kbl() & kable_styling() -> for dataframe object styling
library(plotly) # interactive graphs
library(fmsb) # radarcharts()
library(tidyr) # drop_na() etc.
data <- read.csv("client_class.csv")[-1]
names(data)[37] <- "Client_class"
```

```{r load episode data for program info}
EpisodeDaTa <- read.csv("../Data /EpisodeData.csv")
```

# Summary Statistics

## Summary table of main variables of interests (no demographics)
```{r data prep}
data <- data %>% mutate(Incident_class = ifelse(IncidentScore == 1,"Yes","No"),
                        TraumaHistory = ifelse(TraumaHistory == 1, 1, 0))
data_mat <- data %>% 
## include main variables of interests
  select(Client_class,CurrentAge, SuicideRisk_Total, HomicideRisk_Total, IncidentScore, CriminalHistory, TraumaHistory, AodInvolvement, Disruption, AodLastSixMonths, LegalNonConformingLastSixMonths,GlobalAodPsychosocialDistruptionAndProblems) %>% 
  mutate(Client_class=as.numeric(as.character(Client_class)))
# code out missing values in trauma as 0
data_mat[is.na(data_mat$TraumaHistory),]$TraumaHistory = 0
data_mat[is.na(data_mat$CriminalHistory),]$CriminalHistory = 0
```

```{r create table}
table <- 
  data_mat %>% 
  group_by(Client_class) %>% 
  summarise(`Current Age` = mean(CurrentAge),
            `Suicide Risk Level`=mean(SuicideRisk_Total),
            `Homicide Risk Level`=mean(HomicideRisk_Total),
            `Criminal History` = mean(CriminalHistory),
            `Aod Involvement` = mean(AodInvolvement),
            `Disruption` = mean(Disruption),
            `Life Functioning Problems` = mean(GlobalAodPsychosocialDistruptionAndProblems),
            `Aod Last Six Months` = mean(AodLastSixMonths),
            `Legal Non-Conforming Problems` = mean(LegalNonConformingLastSixMonths))  %>% 
  t() %>% # transpose matrix
  round(digits = 2) %>% 
  as.data.frame() 
table <- table[-1,] # exclude client id
colnames(table) <- c("Class 1", "Class 2", "Class 3")
table2 <- 
      rbind(`Number of Clients`=c(464,469,881),
      table, 
      `Proportion with Trauma History` = c('44% (n=204)','29% (n=134)','32% (n=288)'),
      `Proportion with Incident History` = c('3% (n=12)','5% (n=23)','3% (n=26)'))
```

```{r add table highlights}
# bold highest number of clients
table2[1,3] <- cell_spec(table2[1,3],bold = T)
# bold highest age
table2[2,1] <- cell_spec(table2[2,1],bold = T)

# highlight in suicide risk level
table2[3,1] <- cell_spec(table2[3,1], color = "white", background = "red",bold = T) # highest suicide
table2[3,2] <- cell_spec(table2[3,2], color = "white", background = "green",bold = T) # lowest suicide

# highlight in homicide risk level
table2[4,1] <- cell_spec(table2[4,1], color = "white", background = "red",bold = T) # highest homicide
table2[4,2] <- cell_spec(table2[4,2], color = "white", background = "green",bold = T) # lowest homicide

# highlight in criminal history
table2[5,3] <- cell_spec(table2[5,3], color = "white", background = "red",bold = T) # highest criminal

# highlight in Aod Involvement
table2[6,1] <- cell_spec(table2[6,1], color = "white", background = "red",bold = T) # highest AOD
table2[6,2] <- cell_spec(table2[6,2], color = "white", background = "green",bold = T) # lowest AOD

# highlight in Disruption
table2[7,1] <- cell_spec(table2[7,1], color = "white", background = "red",bold = T) # highest 
table2[7,2] <- cell_spec(table2[7,2], color = "white", background = "green",bold = T) # lowest 

# highlight in Life functioning
table2[8,1] <- cell_spec(table2[8,1], color = "white", background = "red",bold = T) # highest
table2[8,2] <- cell_spec(table2[8,2], color = "white", background = "green",bold = T) # lowest 

# highlight in Aod last 6 months
table2[9,1] <- cell_spec(table2[9,1], color = "white", background = "red",bold = T) # highest AOD
table2[9,2] <- cell_spec(table2[9,2], color = "white", background = "green",bold = T) # lowest AOD

# highlight in legal non-conform
table2[10,1] <- cell_spec(table2[10,1], color = "white", background = "red",bold = T) # highest AOD
table2[10,2] <- cell_spec(table2[10,2], color = "white", background = "green",bold = T) # lowest AOD

# highlight in trauma history
table2[11,1] <- cell_spec(table2[11,1], color = "white", background = "red",bold = T) # highest trauma share

# highlight in Incident history
table2[12,2] <- cell_spec(table2[12,2], color = "white", background = "red",bold = T) # highest incident proportion
```

```{r draw summary table}
# rename table 2
colnames(table2) <- c("Class 1-High", "Class 2-Low", "Class 3-Moderate")
table2 %>% 
  kbl(escape = F, booktabs = T) %>% 
  kable_styling(bootstrap_options = "striped", position = "left")
```

## Complete summary stats of all variables included

```{r summary stats of all classes}
#data[is.na(data$CriminalHistory),]$CriminalHistory = 0
(data %>% 
    select(-c("X2014Rel1Date","State","Zip","AddressType")) %>% 
    mutate(TraumaHistory = as.factor(case_when(is.na(TraumaHistory)~0,TraumaHistory == 1~1,TraumaHistory == 0~0)),
           Gender = as.factor(Gender),
          Race=as.factor(Race),
          PrimaryLanguage = as.factor(PrimaryLanguage),
          UsVeteran = as.factor(UsVeteran),
          ReturnType = as.factor(ReturnType),
          MaritalStatus=as.factor(MaritalStatus),
          Religion=as.factor(Religion),
          Suicide_class = as.factor(Suicide_class),
          Homicide_class = as.factor(Homicide_class),
          Incident_class = as.factor(Incident_class),
          Aod1_class = as.factor(Aod1_class),
          Aod2_class = as.factor(Aod2_class),
          Disruption_class = as.factor(Disruption_class),
          LegalNonConform_class = as.factor(LegalNonConform_class),
          LifeFunctionality_class = as.factor(LifeFunctionality_class),
          `Discharged City`=as.factor(City),
          Client_class = as.factor(as.character(Client_class))) %>% 
  group_by(Client_class) %>% 
  do(summary_stats=summary(.)))$summary_stats
```

# Visualizations
## All Client classes profiles: radarcharts
```{r rescale data}
# get rescaled data matrix of numeric variables using min-max difference normalization
maxmin <- function(x){
  x <- (x-min(x))/(max(x)-min(x))
}
# create rescaled data matrix using max-min difference
data_scaled <- data_mat %>% 
  drop_na() %>% 
  sapply(maxmin) %>% 
  as.data.frame() %>% # convert into dataframe
  mutate(Client_class = data_mat$Client_class) %>% 
  group_by(Client_class) %>% 
  summarise(Age = mean(CurrentAge),
            SuicideRisk_Total=mean(SuicideRisk_Total),
            HomicideRisk_Total=mean(HomicideRisk_Total),
            TraumaHistory = mean(TraumaHistory),
            CriminalHistory = mean(CriminalHistory),
            AodInvolvement = mean(AodInvolvement),
            Disruption = mean(Disruption),
            LifeFunctionProblem = mean(GlobalAodPsychosocialDistruptionAndProblems),
            AodLastSixMonths = mean(AodLastSixMonths),
            LegalNonConform = mean(LegalNonConformingLastSixMonths))
```

```{r cluster 1 profie}
all_cluster1 <- data_scaled %>% 
  filter(Client_class==1) 
# To use the fmsb package, I have to add 2 lines to the dataframe: the max and min of each variable to show on the plot!
all_cluster1 <- rbind(rep(0.75,11) , rep(0,11) , all_cluster1)
radarchart(all_cluster1[-1],axistype=1,
           # custom polygon
           pcol = rgb(1, 0.4, 0.6, 0.9), pfcol =rgb(1, 0.4, 0.6, 0.25), plwd=4,
           # custom grid
           cglcol = "grey", cglty = 1, axislabcol = "grey", caxislabels =  c(0,0.15,0.25,0.5,0.75), cglwd=0.8,
           # custom labels
           vlcex = 0.7, # font size for labels
           vlabels = c("Age","Suicide\nRisk","Homicide\nRisk"," Trauma\nHistory","Criminal\nHistory","Aod\nInvolvement","Disruption","Life\nFunctioning\nProblems","Aod Last\nSix Months","Legal Non-Conforming\nProblems"),
           calcex = 0.8, # font size of center axis labels
           title="Class 1 (High) Profile")
```

```{r cluster 2 profie}
all_cluster2 <- data_scaled %>% 
  filter(Client_class==2) 
# To use the fmsb package, I have to add 2 lines to the dataframe: the max and min of each variable to show on the plot!
all_cluster2 <- rbind(rep(0.75,11) , rep(0,11) , all_cluster2)
radarchart(all_cluster2[-1],axistype=1,
           # custom polygon
           pcol = rgb(0.2,0.5,0.5,0.9), pfcol =rgb(0.2,0.5,0.5,0.5), plwd=4,
           # custom grid
           cglcol = "grey", cglty = 1, axislabcol = "grey", caxislabels = c(0,0.15,0.25,0.5,0.75), cglwd=0.8,
           # custom labels
           vlcex = 0.7, # font size for labels
           vlabels = c("Age","Suicide\nRisk","Homicide\nRisk"," Trauma\nHistory","Criminal\nHistory","Aod\nInvolvement","Disruption","Life\nFunctioning\nProblems","Aod Last\nSix Months","Legal Non-Conforming\nProblems"),
           calcex = 0.8, # font size of center axis labels
           title="Class 2 (Low) Profile")
```

```{r cluster 3 profie}
all_cluster3 <- data_scaled %>% 
  filter(Client_class==3) 
# To use the fmsb package, I have to add 2 lines to the dataframe: the max and min of each variable to show on the plot!
all_cluster3 <- rbind(rep(0.75,11) , rep(0,11) , all_cluster3)
radarchart(all_cluster3[-1],axistype=1,
           # custom polygon
           pcol = 4, pfcol =rgb(0, 0.4, 1, 0.25), plwd=4,
           # custom grid
           cglcol = "grey", cglty = 1, axislabcol = "grey", caxislabels =  c(0,0.15,0.25,0.5,0.75), cglwd=0.8,
           # custom labels
           vlcex = 0.7, # font size for labels
           vlabels = c("Age","Suicide\nRisk","Homicide\nRisk"," Trauma\nHistory","Criminal\nHistory","Aod\nInvolvement","Disruption","Life\nFunctioning\nProblems","Aod Last\nSix Months","Legal Non-Conforming\nProblems"),
           calcex = 0.8, # font size of center axis labels
           title="Class 3 (moderate) Profile")
```

## Client characteristics by class membership
### Demographics by class

```{r class versus age}
data <- mutate(data, class = as.factor(Client_class))
age_plot <- ggplot(data, aes(x=CurrentAge, color=class, fill=class))+
  geom_density(alpha=0.4)+
  theme_minimal()+
  xlab("Current Age")+
  ylab("")+
  ggtitle("Client class by Age")
ggplotly(age_plot)
```

```{r class versus gender}
gender_plot <- data %>% 
  ggplot(aes(x=Gender, fill=Gender))+
  geom_bar()+
  facet_wrap(~class, labeller = label_both, scales="free")+
  ggtitle("Client class by Gender")
ggplotly(gender_plot)
```

```{r class versus race}
race_plot <- 
  ggplot(data,aes(x=Race, fill=Race))+
  geom_bar()+
  facet_grid(rows=vars(class), labeller = label_both, scales="free")+
  ggtitle("Client class by Race")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank() 
        )
ggplotly(race_plot)
```

```{r class versus religion}
religion_plot <- 
  ggplot(data,aes(x=Religion, fill=Religion))+
  geom_bar()+
  facet_grid(rows=vars(class), labeller = label_both, scales="free")+
  ggtitle("Client class by Religion")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank() 
        )
ggplotly(religion_plot)
```

```{r class versus marital status}
marital_plot <- 
  ggplot(data,aes(x=MaritalStatus, fill=MaritalStatus))+
  geom_bar()+
  facet_grid(rows=vars(class), labeller = label_both, scales="free")+
  ggtitle("Client class by Marital Status")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank() 
        )
ggplotly(marital_plot)
```

### ASUS scale factors vs. class

```{r class vs aod1}
aod1_plot <- ggplot(data, aes(x=AodInvolvement, color=class, fill=class))+
  geom_density(alpha=0.4)+
  theme_minimal()+
  xlab("Lifetime AOD Involvement Score")+
  ylab("")+
  ggtitle("Distribution of AOD Lifetime Involvement by Client Class")
ggplotly(aod1_plot)
```

```{r aod last 6 months}
aod2_plot <- data %>% 
  mutate(Aod2_class = factor(Aod2_class, levels=c("low","medium","high"))) %>% 
  ggplot(aes(x=Aod2_class,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("AOD Involvement Level Last Six Months")+
  ylab("")+
  ggtitle("AOD Involvement Level Last Six Months")
ggplotly(aod2_plot)
```

```{r legal non conform}
legal_plot <- data %>% 
  mutate(LegalNonConform_class = factor(LegalNonConform_class, levels=c("low","medium","high"))) %>% 
  ggplot(aes(x=LegalNonConform_class,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Legal Non Conforming Problems Last Six Months")+
  ylab("")+
  ggtitle("Legal Non Conforming Problems")
ggplotly(legal_plot)
```

```{r class vs. disruption}
dis_plot <- data %>% 
  mutate(Disruption_class = factor(Disruption_class, levels=c("low","medium","high"))) %>% 
  ggplot(aes(x=Disruption_class,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Disruption Score")+
  ylab("")+
  ggtitle("Distribution of Disruption by Client Class")
ggplotly(dis_plot)
```

```{r life functionality vs. class}
func_plot <- 
  ggplot(data, aes(x=GlobalAodPsychosocialDistruptionAndProblems,fill=class,color=class))+
  geom_density(alpha=0.5)+
 # facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Life Functionality Problems")+
  ylab("")+
  ggtitle("Life Functionality Problems by Client Class")
ggplotly(func_plot)
```

### Suicide & Homicide risks vs. class

```{r suicide risk}
suicide_plot <- data %>% 
  filter(SuicideRisk_Total != 0) %>% 
  ggplot(aes(x=SuicideRisk_Total,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Suicide Risks Level")+
  ylab("")+
  ggtitle("Suicide Risk Levels by Client Class")
ggplotly(suicide_plot)
```

```{r suicide class risk}
suicide_class_plot <- data %>% 
  ggplot(aes(x=Suicide_class,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Suicide Risks Level")+
  ylab("")+
  ggtitle("Suicide Risk Class by Client Class")
ggplotly(suicide_class_plot)
```

```{r homicide risk}
homicide_plot <- data %>% 
  filter(HomicideRisk_Total != 0) %>% 
  ggplot(aes(x=HomicideRisk_Total,fill=class))+
  geom_bar(alpha=0.7)+
  facet_grid(rows=vars(class),labeller = label_both,scales="free")+
  theme_minimal()+
  xlab("Homicide Risks Level")+
  ylab("")+
  ggtitle("Homicide Risk Levels by Client Class")
ggplotly(homicide_plot)
```

### Trauma history vs. class

```{r trauma history}
trauma_plot <- data %>% 
  ggplot(aes(x=Trauma_class, fill=Trauma_class))+
  geom_bar(alpha=0.8)+
  facet_wrap(~class, labeller = label_both, scales="free")+
  ylab("Trauma History")+
  ggtitle("Client Class vs. Trauma History")
ggplotly(trauma_plot)
```

### Incident scores vs. class

```{r incidents}
incident_plot <- data %>% 
  filter(IncidentScore != 0) %>% 
  ggplot(aes(x=IncidentScore,color=class, fill=class))+
  geom_density(alpha=0.5)+
  facet_grid(labeller = label_both)+
  theme_minimal()+
  xlab("Incident Scores")+
  ylab("")+
  ggtitle("Incident Scores by client class")
ggplotly(incident_plot)
```


# Chi-square Test 

## Chi-square on Incident_class vs. Class membership
```{r chi-square on incident}
chi <- chisq.test(data$Incident_class, data$class)
chi
```

```{r observed cell counts & prop, include=FALSE}
chi$observed # cell counts
prop.table(chi$observed, 2) # col perc --> col 100%
prop.table(chi$observed, 1) # row perc --> row 100%
```

```{r incident history vs. client class}
# group by incident history and class
incident_data <- data %>% 
  group_by(Incident_class, Client_class) %>% 
  summarise(count=n(), .groups="keep") %>% 
  tidyr::spread(Client_class, count) 
names(incident_data) <- c("Incident History","Class 1","Class 2", "Class 3")

# get prop and freq
incident_table <- incident_data %>% 
  mutate(class1_prop = `Class 1`/464,
         class2_prop = `Class 2`/469,
         class3_prop = `Class 3`/881) %>% 
  mutate(`Class 1 - High (n=464)` = paste0(round(class1_prop*100,2), "%"),
         `Class 2 - Low (n=469)` = paste0(round(class2_prop*100,2), "%"),
         `Class 3 - Moderate (n=881)` = paste0(round(class3_prop*100,2), "%")) %>% 
  select(`Incident History`,`Class 1 - High (n=464)`,`Class 2 - Low (n=469)`,`Class 3 - Moderate (n=881)`)
```

```{r highlight incident table}
# highlight row max and min
incident_table[1,2] <- cell_spec(incident_table[1,2], color = "white", background = "green",bold = T) 
incident_table[2,3] <- cell_spec(incident_table[2,3], color = "white", background = "red",bold = T) 
```

```{r draw incident table}
incident_table %>% 
  kbl(escape = F, booktabs = T) %>% 
  kable_styling(bootstrap_options = "striped", position = "left")
```

## Chi-square on Discharged City vs. Class membership
```{r post-hoc}
source('https://raw.githubusercontent.com/PassionDrivenStatistics/R/master/ChiSquarePostHoc.R')

chisq.post.hoc(chi$observed, popsInRows=FALSE, control="bonferroni")
```

```{r chi square location}
chi2 <- chisq.test(data$City, data$Client_class)
chi2
```

```{r location vs. client class}
# group by city and class
city_data <- data %>% 
  group_by(City, Client_class) %>% 
  summarise(count=n(), .groups="keep") %>% 
  tidyr::spread(Client_class, count) 
names(city_data) <- c("Discharged City","Class 1","Class 2", "Class 3")

# get prop and freq
city_table <- city_data %>% 
  arrange(desc(`Class 1`)) %>% 
  tidyr::drop_na() %>% 
  mutate(class1_prop = `Class 1`/464,
         class2_prop = `Class 2`/469,
         class3_prop = `Class 3`/881) %>% 
  mutate(`Class 1 - High (n=464)` = paste0(round(class1_prop*100,2), "%"),
         `Class 2 - Low (n=469)` = paste0(round(class2_prop*100,2), "%"),
         `Class 3 - Moderate (n=881)` = paste0(round(class3_prop*100,2), "%")) %>% 
  select(`Discharged City`,`Class 1 - High (n=464)`,`Class 2 - Low (n=469)`,`Class 3 - Moderate (n=881)`)
```

```{r highlight city table}
# bolding highest row perc
city_table[1,3] <- cell_spec(city_table[1,3], bold = T, color =  "white", background = "orange")
city_table[2,2] <- cell_spec(city_table[2,2], bold = T, color = "white", background = "orange")
city_table[3,4] <- cell_spec(city_table[3,4], bold = T, color = "white", background = "orange")
```

```{r draw top 10 city table}
head(city_table,10) %>% 
  kbl(escape = F, booktabs = T) %>% 
  kable_styling(bootstrap_options = "striped", position = "left")
```

